---
title: SOP 1 Download and calculate forest health metrics within the APPA
  HUC10 Shell
output:
  html_document:
    highlight: textmate
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
version: Built with R `r getRversion()`
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#at<-rFIA::readFIA("C:/Users/hstan/Dropbox/FIA/VT")
#at_FIAMR <- rFIA::clipFIA(at, mask = NULL, matchEval= FALSE, mostRecent = TRUE, nCores = cores)

```
# Introduction {#intro}

The U.S. Forest Service (USFS) administers the Forest Inventory and Analysis Program to acquire inventory data on U.S. forests. The program was launched following the McSweeney - McNary Forest Research Act of 1928 (P.L. 70-466) which led to the first forests being inventoried in 1930 (USFS 2012). Data collected by the program are valuable for a variety of purposes, including timber harvest and ecological health. According to the FIA website, “…FIA reports on status and trends in forest area and location; in the species, size, and health of trees; in total tree growth, mortality, and removals by   harvest; in wood production and utilization rates by various products; and in forest land ownership…”FIA data are organized by state, but all state datasets conform to a consistent format1. However, in order to use data from the APPA region, data must be re-aggregated to the APPA area of interest. 

This document describes how to use the `rFIA` R package ([Stanke et al. 2020](https://doi.org/10.1016/j.envsoft.2020.104664)) to download USFS FIA data and estimate forest attributes described in the APPA Forest Health Protocol ([Dieffenbach 2018](https://irma.nps.gov/DataStore/Reference/Profile/2257434)). The FIA program conducts annual (panel) inventories within each state. For the APPA region, this is most often a series of 5 annual, spatially unbiased inventories within each sampling cycle. This panel structure allows the FIA program to improve the precision of status and change estimates by leveraging previously collected data within an inventory cycle (e.g., estimate for 2015 may include data from annual inventories conducted from 2010-2015).

This document illustrates how to use the functions in `rFIA` to:

1. Access FIA data
2. Subset the FIA data frame to the APPA HUC10 Shell
3. Produce the following estimates, typically at the plot- and ecoregion-scale: 

      * **Live tree abundance**
          + TPA, BAA, biomass, and carbon by species
      * **Species diversity of live trees**
          + Shannon's diversity, evenness, and richness
      * **Tree vital rates**
          + Annual diameter, basal area, and biomass growth by species
      * **Forest demographic rates**
          + Annual recruitment, mortality, and harvest totals and rates by species
      * **Regeneration abundance**
          + TPA of regenerating stems (<5" DBH) by species and size-class
      * **Snag abundance**
          + TPA, BAA, biomass, carbon, relative fraction
      * **Down woody debris abundance**
          + Volume, biomass, and carbon by fuel class
      * **Invasive Plant abundance**
          + % cover by species
      * **Stand structural stage distributions**
          + % area in pole, mature, and late stage forest

<br>

This document does not describe the FIA database structure or demonstrate all of the functionality of the `rFIA` package functions. Readers should consult the documentation in `rFIA` for those details. Formatting of this document follows standards [RMarkdown](https://rmarkdown.rstudio.com/) notation. For instance, examples of executable R code chunks are denoted in gray areas in the following sections.

<br>

# Load packages
``` {r load, include = TRUE, echo= TRUE, eval= FALSE, cache= TRUE,warning=FALSE, comment= NA}
#library(rFIA) # package needs a few updates to fix recent issues
# using Hunter's dev version with personal edits (including accepting a pull request) 
devtools::load_all('C:/Users/JJGross/Documents/R_projects/rFIA-master/rFIA-master/R')
#library (rgdal) # rgdal retired
library(sf) # use sf instead of rgdal
library(parallel)

```

You can access the most recent `rFIA` package release with this command: 
`devtools::install_github("hunter-stanke/rFIA")`

# Setup parallel processing
``` {r para, include = TRUE, echo= TRUE, eval= FALSE, cache= TRUE,warning=FALSE, comment= NA}
## How many physical cores do you have?
parallel::detectCores(logical = FALSE) 

## How many do you want to use?

## If you still want to use your computer for anything else during the computations (e.g., checking your E-mails or writing a Word document), you should reserve at least one core for those remaining tasks.

cores <- parallel::detectCores(logical = FALSE)-1 # set to use all physical cores minus one.
```
 
# Import FIA data for each state intersecting APPA
`getFIA` downloads FIA Data from the FIA Datamart, loads the data into R environment, and provides an option to export the tables. 

This example saves the tables to the directory `'./FIA'` from the 13 states along APPA

``` {r download, include = TRUE, echo= TRUE, eval= FALSE,warning=FALSE, comment= NA,}

## APPA States
at_states <- c('CT', 'GA', 'ME', 'MD', 'MA', 'NH', 'NJ', 'NY', 'NC', 'PA', 
               'TN', 'VT', 'VA')

## Small subset of FIA dataset to test 'getFIA()' before big download
test <- c('AS', 'GU') # smallest datasets in 2023

#FIA_data_here <- './FIA/'
FIA_data_here <- 'C:/Users/JJGross/Documents/R_projects/FIA_data/allStates'

## change between 'states = test' (to test) and ' states = at_states' (for APPA download) 
## 'load = FALSE' downloads the data but does not load data into R session 
## (loading into R session shown in next step with 'readFIA()')

#getFIA(states = test, dir = FIA_data_here, 
#       nCores = cores, common = TRUE, load = FALSE) 
```

This is a large amount of data so this step will be slow. Test dataset useful to check process first. You can choose to load in the downloaded data directly into the R environment with the argument `r load = TRUE`, however, since this is a large database if all data is too large to fit in memory, use `r load = FALSE` and load the data as a "Remote.FIA.Database" with `r `readFIA` as demonstrated in the next step.

# Load FIA database into R environment from local directory

``` {r read, include = TRUE, echo= TRUE, eval= FALSE, warning=FALSE, comment= NA}
at <- readFIA(dir = FIA_data_here, states = at_states, nCores = cores)
```

# Load HUC10 shell polygons

Load in a file containing polygons of ecological sections and subsections within Subregions within the conterminous United States clipped to the HUC10 shell surrounding the Appalachian National Scenic Trail. This data set contains regional geographic delineations for analysis of ecological relationships across ecological units. Currently, all spatial data for this protocol are located on NETN's server.

Load in the data using `sf::read_sf`

``` {r eco, include = TRUE, echo= TRUE, eval= FALSE,warning=FALSE, comment= NA}
#eco <- readOGR('./ecoregions', 'at_ecoSub') #rgdal 
eco <- sf::read_sf(dsn = './ecoregions', layer = 'at_ecoSub')
plot(eco["SUBSECTION"])
plot(eco["SECTION_NA"])

# Distinct SUBSECTIONS:
n_distinct(eco$SUBSECTION)

# Distinct SUBSECTI_1:
n_distinct(eco$SUBSECTI_1)
levels(as.factor(eco$SUBSECTI_1))

# Distinct SECTION_NA:
n_distinct(eco$SECTION_NA)
levels(as.factor(eco$SECTION_NA))

# Distinct PROVINCE_N:
n_distinct(eco$PROVINCE_N)
levels(as.factor(eco$PROVINCE_N))

```

In this case looks in the subdirectory '/ecoregions' of project folder for the layer named 'at_ecoSub'

<br>

# Spatially subset FIA data to HUC10 Shell
`clipFIA` performs space-time queries on Forest Inventory and Analysis Database (FIADB). `clipFIA` subsets the database to include only data associated with particular inventory years (i.e., most recent), and/or only data within a user-defined region.

Spatially, the `mask` argument below selects only the `at` FIA plots which fall within the perimeter of the `eco` spatial extent (i.e. the outer boundary of the HUC10 Shell polygons). 

Temporally, the `mostRecent` argument returns only data for most recent inventory if set to `TRUE`

Finally, by setting `matchEval = TRUE` the `clipFIA` function will return only population evaluations (found in the `POP_EVAL` table) that have a matching end inventory year `END_INVYR`. The `END_INVYR` matches must occur throughout all states in the supplied dataset, any evaluation IDs with unmatched `END_INVYR` are dropped. 



<br>

## Access all inventories in FIADB

```{r intersect, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, , eval= FALSE}
at_FIA <- clipFIA(at, mask = eco, matchEval= FALSE, mostRecent = FALSE, nCores = cores)
```

<br>

## Access the most recent inventory in FIADB

Set `mostRecent = TRUE`

```{r intersect1, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, , eval= FALSE}
at_FIAMR <- clipFIA(at, mask = NULL, matchEval= FALSE, mostRecent = TRUE, nCores = cores)
``` 

<br>

## Access inventories from matching reporting years in FIADB

Set `matchEval = TRUE`

```{r intersect2, include = TRUE, echo=TRUE, warning=FALSE, comment= NA,  eval= FALSE}
at_FIAMatch <- clipFIA(at, mask = eco, matchEval= TRUE, mostRecent = FALSE, nCores = cores)
```

<br>

The following examples will use the `at_FIA` object, which includes all inventories within the HUC10 Shell.

<br>

# Data Archiving and Population Estimation of forest health metrics

Accessing FIA data is needed for archiving raw and summary data and to calculate a vareity of forest health metrics. 

The following section illustrates how to:

1. Create raw data tables (typically sub plot-level) within the APPA HUC10 Shell for archiving with protocol products. 

2. Derive population estimates within the APPA HUC10 Shell at the subplot-, plot- and ecoregion-scale for each forest health metric:

      * **Live tree abundance**
          + TPA, BAA, biomass, and carbon by species
      * **Species diversity of live trees**
          + Shannon's diversity, evenness, and richness
      * **Tree vital rates**
          + Annual diameter, basal area, and biomass growth by species
      * **Forest demographic rates**
          + Annual recruitment, mortality, and harvest totals and rates by species
      * **Regeneration abundance**
          + TPA of regenerating stems (<5" DBH) by species and size-class
      * **Snag abundance**
          + TPA, BAA, biomass, carbon, relative fraction
      * **Down woody debris abundance**
          + Volume, biomass, and carbon by fuel class
      * **Invasive Plant abundance**
          + % cover by species
      * **Stand structural stage distributions**
          + % area in pole, mature, and late stage forest
  
Each of the following subsections illustrates how to use the variety of functions in `rFIA` to calculate these metrics and provides a description of their tabular outputs.


**Population estimation procedures:**

Functions in `rFIA` can be used to derive population estimates of forest data using 5 unique estimators (methods for panel combination) using the `method` argument in each function. In the below examples only the annual method is used (no panel combination). 

* __Temporally Indifferent__ (`method = TI`): The temporally indifferent method assumes that all annual panels within an inventory cycle were collected simultaneously in the reporting year. This is the flagship method used by the FIA program, seen tools like `EVALIDator`, though it introduces temporal lag bias and smoothing that limits its utility for change detection.

* __Annual__ (`method= ANNUAL`): The annual method returns estimates produced directly from annual panels (no panel combination). Thus only data that is measured in a given year is used to produce estimates for that year. This method will produce estimates with higher variance (loss of precision) as it forgoes the opportunity to leverage previous information (increase sample size), although it may be the best option for assessing inter-annual variation and temporal trends. NOTE: This estimator may produce estimates with temporally cyclical structure becuase of repeat observations of individual annual panels between inventory cycles. For example if plots are measured on 5 year cycles beginning in 2005, the same plots will be remeasured in 2010, 2015, 2020 and onward. Hence, corresponding annual panel estimates are likely to be similar to one another, potentially producing results with some odd temporal structure. 

* __Simple Moving Average__ (`method = SMA`): The simple moving average estimator (SMA) is similar to the TI, but rather than treating all plots in an inventory cycle as one large periodic inventory, the SMA instead produces estimates from annual panels and subsequently combines those panels with equal weight. Hence, the SMA is in a sense “temporally-indifferent” by weighting recent information equal to old information (think annual panels here), and represents the average state of the population over the temporal frame of sampling (length of an inventory cylce). A key difference between the SMA and TI: the SMA effectively takes into account the sample size of annual panels, of particular importance for the associated variance estimator.

* __Linear Moving Average__ (`method= LMA`): The linear moving average estimator (LMA) falls in the middle ground between the SMA and annual estimators, leveraging all annual panels within an inventory cycle to produce estimates, although giving higher weight to more recent panels. The relative weights applied to each panel declines linearly as a function of time since measurement. This reduces the effective sample size of the inventory, and hence the LMA tends to be less precise than the SMA (i.e., higher variance). However, a temporally decaying weighting scheme means the LMA tends to have higher temporal specificity than the SMA (less temporal smoothing).

* __Exponential Moving Average__ (`method = EMA`): The exponential moving average estimator (EMA) is the most flexible design-based estimator implemented in rFIA. The EMA is similar to the LMA in that it gives higher weight to more recent panels, however these weights decline exponentially as a function of time since measurement. We can control the rate of this decay with a parameter called `lambda` (0.5 by default, ranges form 0-1). Low lambda values will place higher weight on more recent observations, so as lambda approaches 0, the EMA will approach the annual estimator. In contrast, high lambda values spread weights more evenly across the inventory cycle, so as lambda approaches 1, the EMA will approach the SMA. We face the same trade-off here as we do in the LMA: precision vs. temporal specificity. That is, as lambda increases, we gain precision (reduce variance) and lose temporal specificy (increase temporal smoothing). 

The following example code produces estimates for all FIA reporting years within the APPA HUC10 Shell which were available in each state. Some states have different reporting schedules, and thus there is some missingness in the data. FIA data may not be available for each ecoregion subsection defined by the `at_ecoSub.shp` shapefile. Attributes of ecoregions where FIA data was unavailable are  preserved in objects although estimated attributes are listed as `NA` for these rows. 

<br>

## Stand Structure
`standStruct` estimates  forest structural stage distributions as percent forested land area in pole, mature, late, and mosaic stages and returns the stand structural stage distribution of an area of forest/timberland from FIA data.

### **Plot-level**

Access plot-level structure by:

``` {r raw_ss,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
ssPlot <- standStruct(at_FIA, byPlot= TRUE, 
                      grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), 
                      method = "annual")

ssPlot_nogrpByPLOT <- standStruct(at_FIA, byPlot= TRUE, 
                      grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS), 
                      method = "annual")

ssPlot2 <- standStruct(at_FIA, byPlot= TRUE, 
                      #grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), 
                      method = "annual")
```
Note that the `ECOSUBCD` column comes from the original FIA data, not the at_ecoSub.shp

### **Ecoregion-level** 

To return stand structural estimates by Ecoregions intersecting APPA HUC10 Shell:

``` {r eco_ss,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
ss <- standStruct(at_FIA, polys = eco, nCores = cores, method = 'annual')
```

### **Data Definitions**

* **Description:** Estimates of forest structural stage distributions as percent forested land area in pole, mature, late, and mosaic stages
  + <ins>Diameter Classes:</ins>
    + *Pole*: 11 - 25.9 cm
    + *Mature*: 26 - 45.9 cm
    + *Large*: 46+ cm
  + <ins>Structural Stage Classification:</ins>
    + *Pole Stage*: > 67% BA in pole and mature classes, with more BA in pole than mature.
    + *Mature Stage*: > 67% BA in pole and mature classes, with more BA in mature than pole OR > 67% BA in mature and large classes, with more BA in mature.
    + *Late-Successional Stage:*: > 67% BA in mature and large classes, with more in large.
    + *Mosiac:*: Any plot not meeting above criteria.

* **Data fields:** 
    + *YEAR*: FIA reporting year for current abundance estimates. When `byPlot = FALSE` (i.e., population estimates are returned), the `YEAR` column in the resulting dataframe indicates the final year of the inventory cycle that estimates are produced for. For example, an estimate of current forest area (e.g., 2018) may draw on data collected from 2008-2018, and `YEAR` will be listed as 2018 (consistent with EVALIDator). However, when `byPlot = TRUE` (i.e., plot-level estimates returned), the `YEAR` column denotes the year that each plot was measured (`MEASYEAR`), which may differ slightly from its associated inventory year (`INVYR`).
    + *COVER_PCT*: The estimated percentage of the given `STAGE` class for the area of consideration. Percentage is based on the total forest/timberland in the area of consideration. `STAGE` categories include: POLE, MATURE, LATE, MOSAIC.
    + *nPlots*: number of non-zero plots used to compute current area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

<br>

## Diversity
`diversity` produces estimates of diversity from FIA data. Returns Shannon's Index (H), Shannon's Equitability (Eh), and Richness (S) for alpha (mean/SE of stands), beta, and gamma diversity. Default behavior estimates species diversity, using TPA as a state variable and Species Code (SPCD) to groups of individuals. 

### **Plot-level**

To estimate plot-level Shannon's index, Shannon's equitability, and richness for alpha (mean/SE of stands), beta, and gamma diversity:


``` {r raw_div,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
divPlot <- diversity(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),  method = "annual")
```

### **Ecoregion-level**

To estimate diversity metrics from trees with DBH > 10 in. for each inventory year by Ecoregion intersecting APPA HUC10 Shell:

``` {r eco_div,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
div <- diversity(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')

```


### **Data Definitions**

* **Description:** Estimates of species diversity indicies for live trees (DBH >= 12.7 cm) at alpha, beta, and gamma levels.

* **Data fields:** 
    + *YEAR*: FIA reporting year for current abundance estimates
    + *H_a*: estimate of Shannon's Diversity Index for live trees at the alpha (stand) level
    + *H_b*: estimate of Shannon's Diversity Index for live trees at the beta (landscape) level
    + *H_g*: estimate of Shannon's Diversity Index for live trees at the gamma (regional) level    
    + *H_a*: estimate of Shannon's Equitability Index for live trees at the alpha (stand) level
    + *H_b*: estimate of Shannon's Equitability Index for live trees at the beta (landscape) level
    + *H_g*: estimate of Shannon's Equitability Index for live trees at the gamma (regional) level  
    + *S_a*: estimates of species richness for live trees at the alpha (stand) level
    + *S_b*: estimates of species richness for live trees at the beta (landscape) level
    + *S_g*: estimates of species richness for live trees at the gamma (regional) level 
    + *nStands*: number of non-zero stands (conditions) used to compute alpha-level diversity estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

<br>

## Trees Per Acre (TPA) and Basal Area
`tpa` produces tree per acre (TPA) and basal area per acre (BAA) estimates.

### **Individual-level**

To return individual tree records per year at the sub-plot level:

``` {r raw_tpa,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
trees<- tpa(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT,SUBP,STATUSCD,TREE, DIA), bySpecies = TRUE, bySizeClass =TRUE, method = "annual")
``` 

This returns a large `data.frame` with each row corresponding to an individual tree. Returns DBH and host status; see File structure for data column descriptions. This `data.frame` can then be exported using any efficient export funtion (`write.csv())`).

### **Plot-level**

To estimate TPA and Biomass of trees (>10 in. DBH) at the plot-level:

``` {r pl_tpa,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
tpaSppPlot <- tpa(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), bySpecies = TRUE,  method = "annual")
```


To calcuate by size class include argument `bySizeClass =TRUE`


### **Ecoregion-level** 

To return estimates of live trees per acre (DBH >10 in.) and basal area by Ecoregions intersecting APPA HUC10 Shell:

``` {r eco_tpa,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
tpaEco <- tpa(at_FIA, polys = eco, bySpecies = TRUE, treeType = 'live', treeDomain = DIA >= 10, nCores = cores, method = 'annual')
```


### **Data definitions**

* **Description:** Estimates of live tree abundance (TPA & BAA) per forested acre 

* **Data fields:**
    + *YEAR*: FIA reporting year for current abundance estimates
    + *TPA*: estimate of trees per acre
    + *BAA*: estimate of tree basal area (sq.ft.) per acre
    + *TPA_PERC*: estimate of proportion of trees which are live, with respect to TPA
    + *BAA_PERC*: estimate of proportion of trees which are live, with respect to BAA
    + *nPlots_TREE*: number of non-zero plots used to compute tree and basal area estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Regeneration (Stems > 1 in. DBH)
`tpa` can also be used to estimate regeneration by setting the `treeDomain` argument to an appropriate size class distribution.

### **Plot-level**

To estimate regeneration per species (<5 in. DBH) at the plot-level

``` {r pl_regen,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
regenS <- tpa(at_FIA, polys=eco, byPlot = TRUE, bySpecies = TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),treeType = 'live', treeDomain = DIA < 5, nCores = cores, method = 'annual')
```

### **Ecoregion-level**

To estimate regeneration per species (<5 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_regen,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
regen <- tpa(at_FIA, polys = eco, treeType = 'live', treeDomain = DIA < 5, nCores = cores, method = 'annual')
```

<br>

### **Data Definitions**

* **Description:** Estimates of live sapling (2.5 cm <= DBH < 12.7 cm) abundance (TPA & BAA) per forested acre 

* **Data fields:**
    + *YEAR*: FIA reporting year for current abundance estimates
    + *TPA*: estimate of saplings per acre
    + *BAA*: estimate of sapling basal area (sq.ft.) per acre
    + *TPA_PERC*: estimate of proportion of saplings which are live, with respect to TPA
    + *BAA_PERC*: estimate of proportion of saplings which are live, with respect to BAA
    + *nPlots_TREE*: number of non-zero plots used to compute tree and basal area estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

## Mortality and Recruitment

`growMort` produces estimates of annual recruitment, natural mortality, and harvest rates.

Recruitment events are defined as when a live stem which is less than 5 inches DBH at time 1, grows to or beyond 5 inches DBH by time 2. This does NOT include stems which grow beyond the 5-inch diameter criteria and are then subject to mortality prior to remeasurement. Natural mortality is defined as when a live stem is subject to non-harvest mortality between successive measurement periods. Finally, harvest is defined as when a live stem is cut and removed between successive measurements.

### **Plot-level**

To estimate plot-level mortality per species (trees DBH > 10 in.):

``` {r pl_gMort,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
gmSPlot <- growMort(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),  method = "annual", treeDomain = DIA >= 10)

## Recruitment estimates will be null here, as we are excluding stems < 10 inches
```

To include recruitment estimates, remove size restriction in `treeDomain`:

```{r pl_gMort2,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
gmSPlot_recr <- growMort(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),  method = "annual")

#This will also return mortality estimates as well. 
```

To calculate by size class include argument `bySizeClass = TRUE`


### **Ecoregion-level**

To estimate recruitment and mortality (trees DBH > 10 in.) per species by Ecoregions intersecting APPA HUC10 Shell:

``` {r eco_gMort,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
## Same as above, no recruitment estimates because we are excluding stems < 10 inches
gm <- growMort(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')

## To include recruitment
gm_recr <- growMort(at_FIA, polys = eco, nCores = cores, method = 'annual')
``` 

### **Data Definitions**

* **Description:** Estimates of annual tree recruitment, mortality, and harvest rates per forested acre and relative to population totals (e.g. % mortality / year)

* **Data fields:**
    + *YEAR*: FIA reporting year for growth, removal, and mortality estimates
    + *RECR_TPA*: estimate of annual recruitment (growth beyond 10 in. DBH) as trees per forested acre
    + *MORT_TPA*: estimate of annual mortality (excluding harvest and conversion) as trees per forested acre
    + *REMV_TPA*: estimate of annual removals (harvest and conversion) as trees per forested acre
    + *RECR_PERC*: estimate of annual recruitment rate, as % of individuals recruiting relative to total population
    + *MORT_PERC*: estimate of annual mortality rate, as % of individuals subject to mortality relative to total population
    + *REMV_PERC*: estimate of annual removal rate, as % of individuals subject to removal relative to total population
    + *nPlots_TREE*: number of non-zero plots used to compute total tree estimates
    + *nPlots_RECR*: number of non-zero plots used to compute recruitment estimates  
    + *nPlots_MORT*: number of non-zero plots used to compute mortality estimates    
    + *nPlots_REMV*: number of non-zero plots used to compute removal estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    

<br>

## Growth rates
`vitalRates` provides estimates of average annual DBH (inches/ yr), basal area (sq. ft./ yr), biomass (short tons/ yr), and net volume (cu. ft./yr) growth rates for individual stems, along with average annual basal area and net volume growth per acre.


### **Plot-level**

To estimate rates per species (>10 in. DBH) at the plot-level.

``` {r pl_vit,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
vrS  <- vitalRates(at_FIA, byPlot = TRUE, bySpecies = TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```


To calcuate by size class include argument `bySizeClass =TRUE`


### **Ecoregion-level**

Estimate rates per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_vit,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE }
vr  <- vitalRates(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```

### **Data Definitions**

* **Description:** Estimates of (1) individual live tree diameter, basal area, net volume, and biomass growth rates, and (2) basal area, net volume, and biomass growth rates expressed on a per forested acre basis

* **Data fields:**
    + *YEAR*: FIA reporting year for growth, removal, and mortality estimates
    + *DIA_GROW*: estimate of annual diameter growth (inches) rate for live trees (individual growth)
    + *BA_GROW*: estimate of annual basal area growth (sq.ft.) rate for live trees (individual growth)
    + *NETVOL_GROW*: estimate of annual net volume growth (cu.ft.) rate for live trees (individual growth)
    + *BIO_GROW*: estimate of annual biomass growth (tons) rate for live trees (individual growth)
    + *BAA_GROW*: estimate of annual basal area growth (sq.ft.) per forested acre (stand-level growth)
    + *NETVOL_GROW_AC*: estimate of annual net volume growth (cu.ft.) per forested acre (stand-level growth)
    + *BIO_GROW_AC*: estimate of annual biomass growth (tons) per forested acre (stand-level growth)
    + *nPlots_TREE*: number of non-zero plots used to compute tree estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

<br>

## Biomass
`biomass` produces estimates of volume (cu.ft./acre), biomass (tons/acre), and carbon (tons/acre).

### **Individual-level**

To return biomass metrics per stem

```{r ind_bio,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bioS<-biomass(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT,SUBP,STATUSCD,TREE, DIA), bySpecies = TRUE, bySizeClass =TRUE, method = "annual")
```

### **Plot-level**

To estimate biomass per species (>10 in. DBH) at the plot-level

``` {r pl_bio,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bioS <- biomass(at_FIA, byPLot= TRUE, bySpecies = TRUE,grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```

### **Ecoregion-level**

To estimate biomass per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_bio,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bio <- biomass(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```

### **Data Definitions**

* **Description:** Estimates of live tree biomass, volume, and carbon on a per forested acre basis 

* **Data fields:**
    + *YEAR*: FIA reporting year for current abundance estimates
    + *NETVOL_ACRE*: estimate of live tree net volume (cu.ft.) per acre
    + *SAWVOL_ACRE*: estimate of live tree merchantable saw volume (cu.ft.) per acre
    + *BIO_AG_ACRE*: estimate of live tree aboveground biomass (tons) per acre
    + *BIO_BG_ACRE*: estimate of live tree belowground biomass (tons) per acre
    + *BIO_ACRE*: estimate of live tree total (AG + BG) biomass (tons) per acre
    + *CARB_AG_ACRE*: estimate of live tree aboveground carbon (tons) per acre
    + *CARB_BG_ACRE*: estimate of live tree belowground carbon (tons) per acre
    + *CARB_ACRE*: estimate of live tree total (AG + BG) carbon (tons) per acre
    + *nPlots_VOL*: number of non-zero plots used to compute volume, biomass, and carbon estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Snags (dead trees)
`tpa` and `biomass` can also be used to estimate snag density, volume, and biomass by setting the `treeType` argument to `'dead'`.

### **Plot-level**

To estimate snag density (trees/acre) per species (>10 in. DBH) at the plot-level

``` {r pl_snagD,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagDPlot <- tpa(at_FIA, treeType = 'dead', byPlot = TRUE,  treeDomain = DIA >= 10, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```

To estimate snag volume (cu.ft./acre) and biomass (tons/acre) per species (>10 in. DBH) at the plot-level

```{r pl_snagV,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagVPlot <- biomass(at_FIA, treeType = 'dead', byPlot = TRUE,  treeDomain = DIA >= 10, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```

### **Ecoregion-level**

To estimate snag density (trees/acre) per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_snagD,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagD <- tpa(at_FIA, treeType = 'dead', polys = eco, treeDomain = DIA >= 10, nCores = cores, method = 'annual')
```

To estimate snag volume (cu.ft./acre) and biomass (tons/acre) per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_snagV,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagV <- biomass(at_FIA, treeType = 'dead', polys = eco, treeDomain = DIA >= 10, nCores = cores, method = 'annual')
```

### **Data Definitions**

* **Description:** Estimates of standing dead tree abundance (TPA & BAA) per forested acre and relative to population totals (e.g. percent snags by TPA)

* **Data fields:**
  + **Trees per acre (TPA) and Basal Area(BAA):**
    + *YEAR*: FIA reporting year for current abundance estimates
    + *TPA*: estimate of snags per acre
    + *BAA*: estimate of snags basal area (sq.ft.) per acre
    + *TPA_PERC*: estimate of proportion of snags relative to all stems (live and dead), with respect to TPA
    + *BAA_PERC*: estimate of proportion of snags relative to all stems (live and dead), with respect to BAA
    + *nPlots_TREE*: number of non-zero plots used to compute tree and basal area estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
  + **Biomass**: Estimates of standing dead tree (DBH >= 10 in.) biomass, volume, and carbon on a per forested acre basis 
    + *YEAR*: FIA reporting year for current abundance estimates
    + *NETVOL_ACRE*: estimate of snag net volume (cu.ft.) per acre
    + *SAWVOL_ACRE*: estimate of snag merchantable saw volume (cu.ft.) per acre
    + *BIO_AG_ACRE*: estimate of snag aboveground biomass (tons) per acre
    + *BIO_BG_ACRE*: estimate of snag belowground biomass (tons) per acre
    + *BIO_ACRE*: estimate of snag total (AG + BG) biomass (tons) per acre
    + *CARB_AG_ACRE*: estimate of snag aboveground carbon (tons) per acre
    + *CARB_BG_ACRE*: estimate of snag belowground carbon (tons) per acre
    + *CARB_ACRE*: estimate of snag total (AG + BG) carbon (tons) per acre
    + *nPlots_VOL*: number of non-zero plots used to compute volume, biomass, and carbon estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Down woody debris
`dwm` produces estimates of down woody material volume (cu.ft/acre) and biomass (tons/acre). As defined by FIA, down woody material includes dead organic materials (resulting from plant mortality and leaf turnover) and fuel complexes of live shrubs and herbs. To maintain relevance for forest fuels management, we report estimates grouped by fuel lag-time classes. See Woodall et al. (2007) for definitions of fuel lag-time classes.

### **Plot-level**

To estimate down woody debris volume (cu.ft/acre) and biomass (tons/acre) at the plot-level

```{r pl_dwm,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
dwPlot <- dwm(at_FIA, byPlot = TRUE,grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```


### **Ecoregion-level**

To estimate down woody debris volume (cu.ft/acre) and biomass (tons/acre) by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_dwm, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
dw <- dwm(at_FIA, polys = eco, tidy = FALSE, nCores = cores, method = 'annual')
``` 

### **Data Definitions**
* **Description:** Estimates of down woody material volume, biomass, and carbon by fuel type on a per forested acre basis

* **Data fields:**
    + *YEAR*: FIA reporting year for current abundance estimates
    + *VOL_DUFF_ACRE*: Not available for this region
    + *VOL_LITTER_ACRE*: Not available for this region
    + *VOL_1HR_ACRE*: estimate of 1 HR (small fine woody debris) volume (cu.ft.) per forested acre
    + *VOL_10HR_ACRE*: estimate of 10 HR (medium fine woody debris) volume (cu.ft.) per forested acre
    + *VOL_100HR_ACRE*: estimate of 100 HR (large fine woody debris) volume (cu.ft.) per forested acre
    + *VOL_1000HR_ACRE*: estimate of 1000 HR (coarse woody debris) volume (cu.ft.) per forested acre
    + *VOL_PILE_ACRE*: estimate of slash pile volume (cu.ft.) per forested acre
    + *VOL_ACRE*: estimate of total down woody debris (duff + litter + 1HR + 10HR + 100HR + 1000HR + pile) volume (cu.ft.) per forested acre
    + *BIO_DUFF_ACRE*: estimate of duff biomass (tons) per forested acre
    + *BIO_LITTER_ACRE*: estimate of litter biomass (tons) per forested acre
    + *BIO_1HR_ACRE*: estimate of 1 HR (small fine woody debris) biomass (tons) per forested acre
    + *BIO_10HR_ACRE*: estimate of 10 HR (medium fine woody debris) biomass (tons) per forested acre
    + *BIO_100HR_ACRE*: estimate of 100 HR (large fine woody debris) biomass (tons) per forested acre
    + *BIO_1000HR_ACRE*: estimate of 1000 HR (coarse woody debris) biomass (tons) per forested acre
    + *BIO_PILE_ACRE*: estimate of slash pile biomass (tons) per forested acre
    + *BIO_ACRE*: estimate of total down woody debris (duff + litter + 1HR + 10HR + 100HR + 1000HR + pile) biomass (tons) per forested acre
    + *CARB_DUFF_ACRE*: estimate of duff carbon (tons) per forested acre
    + *CARB_LITTER_ACRE*: estimate of litter carbon (tons) per forested acre
    + *CARB_1HR_ACRE*: estimate of 1 HR (small fine woody debris) carbon (tons) per forested acre
    + *CARB_10HR_ACRE*: estimate of 10 HR (medium fine woody debris) carbon (tons) per forested acre
    + *CARB_100HR_ACRE*: estimate of 100 HR (large fine woody debris) carbon (tons) per forested acre
    + *CARB_1000HR_ACRE*: estimate of 1000 HR (coarse woody debris) carbon (tons) per forested acre
    + *CARB_PILE_ACRE*: estimate of slash pile carbon (tons) per forested acre
    + *CARB_ACRE*: estimate of total down woody debris (duff + litter + 1HR + 10HR + 100HR + 1000HR + pile) carbon (tons) per forested acre
    + *nPlots*: number of non-zero plots used to compute down woody material abundance and area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Invasive plant abundance
`invasive` produces estimates of the areal coverage (%) of invasive species and frequency of plots invasives species were detected in.

Estimates of areal coverage of individual invasive species should NOT be summed to produce estimates of areal coverage by ALL invasive species, as areal coverage by species is not mutually exclusive (multiple species my occur in the same area). Current FIA data collection protocols do not allow for the unbiased estimation of areal coverage by all invasive species.

### **Plot-level**

To estimate invasive plant cover per species at the plot-level

```{r pl_inv,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
invPlot <- invasive(at_FIA, byPlot = TRUE,grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```


### **Ecoregion-level**
While the invasive data are fairly sparse (as of 2020) in the FIADB, you can estimate invasive plant cover per species by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_inv,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
inv <- invasive(at_FIA, polys = eco, nCores = cores, method = 'annual')
```
### **Data Definitions**

* **Description:** Estimates of areal coverage by invasive plant species relative to forested land area

* **Data fields:**
    + *YEAR*: FIA reporting year for current abundance estimates
    + *SYMBOL*: unique species ID from NRCS Plant Reference Guide
    + *SCIENTIFIC_NAME*: latin name of the species
    + *COMMON_NAME*: commmon name of the species
    + *COVER_PCT*: estimate of percent areal coverage with respect to total forested land area
    + *nPlots_INV*: number of non-zero plots used to compute species coverage estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
### Forest Stability Index

``` {r fsi,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE }
fsi.NE<-fsi(at_FIA, polys=eco, returnSpatial = FALSE, bySpecies=TRUE)

fsi.NE<-filter(fsi.NE, YEAR == "2020" & SCIENTIFIC_NAME == "Fraxinus americanus")

fsi.plot<-plotFIA(fsi.NE, y= FSI, x= SUBSECTI_1, plot.title = 'Forest Stability Index', x.lab = "Year", y.lab= "Forest Stability Index", se= FALSE, n.max= 5)

#recode x-variable to not show date
fsi.plot$data$xVar<-as.numeric(fsi.plot$data$YEAR)
fsi.plot




fsi.NE.shp<-fsi(at_FIA, polys=eco, returnSpatial = TRUE, bySpecies=TRUE)

plotFIA(fsi.NE.shp, y= FSI)

```


# Basic visualizations

<br>

## Times series plot

### TPA and Basal area over time
```{r ts_plot,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE, fig.height= 8, fig.width= 6}
tpaEcots <- tpa(at_FIA, polys = eco, treeType = 'live', treeDomain = DIA >= 10, returnSpatial = FALSE, method='SMA')

ts.plot<-plotFIA(tpaEcots, y = BAA, grp = SUBSECTI_1, plot.title = 'Live Basal Area (sq.ft/acre) per ecoregion', x.lab = "Year", y.lab= "Live Basal Area (sq.ft/acre) + SE", se= FALSE, n.max= 5)

#recode x-variable to not show date
ts.plot$data$xVar<-as.numeric(ts.plot$data$YEAR)
ts.plot

plotFIA(tpaEcots, y = TPA, grp = SUBSECTI_1, plot.title = 'TPA per ecoregion')
```

## Size class distribution

### Basal area by size class
```{r SCplot, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, cache=TRUE,fig.height= 8, fig.width= 6, eval= FALSE}
tpaEcoSc <- tpa(at_FIAMR, treeType = 'live',  method='annual', bySizeClass = TRUE)

plotFIA(tpaEcoSc, y = BAA, x= sizeClass, plot.title = 'Basal area (sq.ft/acre)')
```

Plotting object `at_FIAMR` which only includes most recent inventory.

<br>

## Plot locations

### Biomass

```{r bio_plot,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bioPl <- biomass(at_FIAMR, byPlot = TRUE, returnSpatial = TRUE, polys = eco)

plot(bioPl)

plot(bioPl['BIO_ACRE'])# biomass per acre
```

## Chloropleth map

### Live TPA and basal area per ecoregion
```{r tpa_plot,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
tpaEco <- tpa(at_FIA, polys = eco, treeType = 'live', treeDomain = DIA >= 10, returnSpatial = TRUE, method='annual')

plotFIA(tpaEco, y = TPA, legend.title = 'Trees per acre')

plotFIA(tpaEco, y = BAA, legend.title = 'Basal area (sq.ft/acre)')
```

Note the need to pass a spatial object to `polys` argument in `tpa`.

<br>


# References

Bechtold, W.A.; Patterson, P.L., eds. 2005. The Enhanced Forest Inventory and Analysis Program - National Sampling Design and Estimation Procedures. Gen. Tech. Rep. SRS - 80. Asheville, NC: U.S. Department of Agriculture, Forest Service, Southern Research Station. 85 p. https://www.srs.fs.usda.gov/pubs/gtr/gtr_srs080/gtr_srs080.pdf

Dieffenbach, F. (2018). Appalachian National Scenic Trail forest health monitoring protocol, Natural Resource Report NPS/NETN/NRR—2018/1804. National Park Service, Fort Collins, Colorado. https://irma.nps.gov/DataStore/DownloadFile/610353

Stanke, H., Finley, A. O., Weed, A. S., Walters, B. F., & Domke, G. M. (2020). rFIA: An R package for estimation of forest attributes with the US Forest Inventory and Analysis database. Environmental Modelling & Software, 127, 104664. https://doi.org/10.1016/j.envsoft.2020.104664

Woodall, C.; Monleon, V.J., eds. 2007. Sampling Protocol, Estimation, and Analysis Procedures for the Down Woody Materials Indicator of the FIA Program. Gen. Tech. Rep. NRS - 22. Newtown Square, PA: U.S. Department of Agriculture, Forest Service, Northern Research Station. https://www.nrs.fs.fed.us/pubs/gtr/gtr_nrs22.pdf

FIA Database User Guide: https://www.fia.fs.fed.us/library/database-documentation/

# Revision History 

Version numbers will be incremented by a whole number (e.g., Version 1.3 to 2.0) when a change is made that significantly affects requirements or procedures. Version numbers will be incremented by decimals (e.g., Version 1.06 to Version 1.07) when there are minor modifications that do not affect requirements or procedures included in the protocol. Add rows as needed for each change or set of changes tied to an updated version number.

**Revision History Log**

| Version #  |  Date |  Revised by  | Changes  |Justification |
|---|---|---|---|---|
| 1.00  |  July 2017 |  Fred Dieffenbach |  Initial Version |   |
| 2.00  | May 2020  | Aaron Weed  | Updated and merged prior versions of SOPs 1 and 2 into this SOP to incorporate new R-based FIA data workflow utilizing ther `rFIA` package.  | These changes were made because former workflow using MS Access downloads from FIA are no longer supported and because analysis workflow for estimating metrics was updated.  |
|   |   |   |   |   |

