---
title: "Download FIA data tables and calculate metrics within the APPA HUC10 Shell"
author:
date: 
version: "Built with R `r getRversion()`"
output: 
  pdf_document:
    
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
   
   theme: united
   highlight: textmate 
   toc: true
   toc_depth: 3
   toc_float: true
   number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction {#intro}

This document describes how to use the `rFIA` R package to download USFS FIA data and estimate forest attributes described in the APPA Forest Health Protocol ([Dieffenbach 2018](https://irma.nps.gov/DataStore/Reference/Profile/2257434)). The FIA program conducts annual (panel) inventories within each state. For the APPA region, this is most often a series of 5 annual, spatially unbiased inventories within each sampling cycle. This panel structure allows the FIA program to improve the precision of status and change estimates by leveraging previously collected data within an inventory cycle (e.g., estimate for 2015 may include data from annual inventories conducted from 2010-2015).

This document illustrates how to use the functions in `rFIA` to access FIA data, subset the FIA data frame to the APPA HUC10 Shell, and produce the following estimates, typically at the plot- and ecoregion-scale. 

1. **Live tree abundance**
    + TPA, BAA, biomass, and carbon by species
2. **Species diversity of live trees**
    + Shannon's diversity, eveness, and richness
3. **Tree vital rates**
    + Annual diameter, basal area, and biomass growth by species
4. **Forest demographic rates**
    + Annual recruitment, mortality, and harvest totals and rates by species
5. **Regeneration abundance**
    + TPA of regenerating stems (<5" DBH) by species and size-class
6. **Snag abundance**
    + TPA, BAA, biomass, carbon, relative fraction
7. **Down woody debris abundance**
    + Volume, biomass, and carbon by fuel class
8. **Invasive Plant abundance**
    + % cover by species
9. **Stand structural stage distributions**
    + % area in pole, mature, and late stage forest

<br>

This document does not describe the FIA database structure or demonstrate all of the functionality of the `rFIA` package functions. Readers shoud consult the documentation in `rFIA` for those details.

<br>

# Load packages
``` {r load, include = TRUE, echo= TRUE, eval= TRUE, cache= TRUE,warning=FALSE, comment= NA}
library(rFIA)
library (rgdal)
library(parallel)
```

You can access the most recent `rFIA` package release with this command: 
`devtools::install_github("hunter-stanke/rFIA")`

# Setup parallel processing
``` {r para, include = TRUE, echo= TRUE, eval= TRUE, cache= TRUE,warning=FALSE, comment= NA}
## How many cores do you have?
parallel::detectCores(logical = FALSE) 

## How many do you want to use?
 cores <- 1 
```
 
# Import FIA data for each state intersecting APPA
`getFIA` downloads FIA Data from the FIA Datamart, loads the data into R environment, and provides an option to export the tables. 

This example saves the tables to the directory `'./FIA'` from the 13 states along APPA

``` {r download, include = TRUE, echo= TRUE, eval= FALSE,warning=FALSE, comment= NA,}
at <- getFIA(states = c('CT', 'GA', 'ME', 'MD', 'MA', 'NH', 'NJ', 'NY', 'NC', 'PA', 'TN', 'VT', 'VA'), nCores = cores,dir = './FIA/', common = TRUE 
```

This is a large amount of data so this step will be slow. 

<br>

# Load HUC10 shell polygons

Load in a file containing polygons of ecological sections and subsections within Subregions within the conterminous United States clipped to the HUC10 shell surrounding the Appalachian National Scenic Trail. This data set contains regional geographic delineations for analysis of ecological relationships across ecological units.

Load in the data using `rgdal::readOGR`

``` {r eco, include = TRUE, echo= TRUE, eval= FALSE,warning=FALSE, comment= NA}
eco <- readOGR('./ecoregions', 'at_ecoSub')
```

Looking in the subdirectory of your project folder '/ecoregions' for the layer named 'at_ecoSub'

<br>

# Spatially subset FIA data to HUC10 Shell
`clipFIA` performs space-time queries on Forest Inventory and Analysis Database (FIADB). You subset database to include only data associated with particular inventory years (i.e., most recent), and/or only data within a user-defined region.

The below examples spatially subsets the `at` FIA data object to the HUC10 Shell polygon `eco` using the `mask` argument. Additional arguments as shown below can access different inventory years.

<br>

## Access all inventories

```{r intersect, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, , eval= FALSE}
at_FIA <- clipFIA(at, mask = eco, matchEval= FALSE, mostRecent = FALSE, nCores = cores)
```

<br>

## Access the most recent inventory

Set `mostRecent = TRUE`

```{r intersect1, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, , eval= FALSE}
at_FIAMR <- clipFIA(at, mask = NULL, matchEval= FALSE, mostRecent = TRUE, nCores = cores)
``` 

<br>

## Access inventories from matching reporting years

Set `matchEval = TRUE`

```{r intersect2, include = TRUE, echo=TRUE, warning=FALSE, comment= NA,  eval= FALSE}
at_FIAMatch <- clipFIA(at, mask = NULL, matchEval= TRUE, mostRecent = FALSE, nCores = cores)
```

<br>

The following examples will use the `at_FIA` object including all inventories within the HUC10 Shell.

<br>

# Data Archiving and Population Estimation

The following section illustrates how to:

+ Create raw data tables (typically sub plot-level) within the APPA HUC10 Shell for archiving with project deliverables. 

+ Derive Population Estimates within the APPA HUC10 Shell at the plot- and ecoregion-scale:
  
Functions in `rFIA` can be used to derive population estimates of forest data using 3 unique estimators (methods for panel combination). In the below examples only the annual method is used (no panel combination). 

* __Temporally Indifferent__ : The temporally indifferent method assumes that all annual panels within an inventory cycle were collected simultaneously in the reporting year. This is the flagship method used by the FIA program, seen tools like `EVALIDator`, though it introduces temporal lag bias and smoothing that limits its utility for change detection.

* __Annual__ : The annual method returns estimates produced directly from annual panels (no panel combination). Thus only data that is measured in a given year is used to produce estimates for that year. This method will produce estimates with higher variance (loss of precision) as we forgo the opportunity to leverage previous information (increase sample size), although it may be the best option for assessing inter-annual variation and temporal trends. NOTE: This estimator may produce estimates with temporally cyclical structure becuase of repeat observations of individual annual panels between inventory cycles. For example if plots are measured on 5 year cycles beginning in 2005, the same plots measured in 2005 will be remeasured in 2010, 2015, 2020 and onward. Hence, corresponding annual panel estimates are likely to be similar to one another, potentially producing results with some odd temporal structure. 

* __Exponential Moving Average__ : The exponential moving average method falls in the middle ground between the temorally indifferent and annual estimators, leveraging all annual panels within an inventory cycle to produce estimates, although weighting recent panels higher than less recent panels. The relative weights applied to each panel declines exponentially as a function of time since measurement, but we can control the rate of this decay with a parameter called `lambda` (ranges form 0-1).  Low lambda values will place higher weight on more recent observations, so as lambda approaches 0, the exponential moving average will approach the annual method for panel combination. For this analysis we used the defualt value of 0.50. 

The following example code produces estimates for all FIA reporting years within the APPA HUC10 Shell which were available in each state. Some states have different reporting schedules, and thus there is some missingness in the data. FIA data may not be available for each ecoregion subsection defined by the `at_ecoSub.shp` shapefile. Attributes of ecoregions where FIA data was unavailable are  preserved in objects although estimated attributes are listed as `NA` for these rows. 

<br>

## Stand Structure
`standStruct` returns the stand structural stage distribution of an area of forest/timberland from FIA data.

### **Plot-level**

Access plot-level structure by:

``` {r raw_ss,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
ssPlot <- standStruct(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),
                       method = "annual")
```

### **Ecoregion-level** 

To return stand structural estimates by Ecoregions intersecting APPA HUC10 Shell:

``` {r eco_ss,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
ss <- standStruct(at_FIA, polys = eco, tidy = FALSE, nCores = cores, method = 'annual')
```

### **Data Definitions**

* **Stand Structure**: Estimates of forest structural stage distributions as percent forested land area in pole, mature, late, and mosaic stages
    + *YEAR*: FIA reporting year for current abundance estimates
    + *POLE_PERC*: estimate of proportion forested area in pole stage forest
    + *MATURE_PERC*: estimate of proportion forested area in mature stage forest
    + *LATE_PERC*: estimate of proportion forested area in late-seral stage forest
    + *MOSAIC_PERC*: estimate of proportion forested area in mosaic stage (unclassified) forest
    + *nPlots*: number of non-zero plots used to compute current area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `ecoregions/at_ecoSub.shp`

<br>

## Diversity
`diversity` produces estimates of diversity from FIA data. Returns Shannon's Index (H), Shannon's equitability (Eh), and richness (S) for alpha (mean/SE of stands), beta, and gamma diversity. Default behavior estimates species diversity, using TPA as a state variable and Species Code (SPCD) to groups of individuals. 

### **Plot-level**

To estimate plot-level Shannon's index, Shannon's equitability, and richness for alpha (mean/SE of stands), beta, and gamma diversity:


``` {r raw_div,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
divPlot <- diversity(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),  method = "annual")
```

### **Ecoregion-level**

To estimate diversity metrics from trees with DBH > 10 in. for each inventory year by Ecoregion intersecting APPA HUC10 Shell:

``` {r eco_div,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
div <- diversity(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')

```


### **Data Definitions**

* **Diversity** Estimates of species diversity indicies for live trees (DBH >= 12.7 cm) at alpha, beta, and gamma levels
    + *YEAR*: FIA reporting year for current abundance estimates
    + *H_a*: estimate of Shannon's Diversity Index for live trees at the alpha (stand) level
    + *H_b*: estimate of Shannon's Diversity Index for live trees at the beta (landscape) level
    + *H_g*: estimate of Shannon's Diversity Index for live trees at the gamma (regional) level    
    + *H_a*: estimate of Shannon's Equitability Index for live trees at the alpha (stand) level
    + *H_b*: estimate of Shannon's Equitability Index for live trees at the beta (landscape) level
    + *H_g*: estimate of Shannon's Equitability Index for live trees at the gamma (regional) level  
    + *S_a*: estimates of species richness for live trees at the alpha (stand) level
    + *S_b*: estimates of species richness for live trees at the beta (landscape) level
    + *S_g*: estimates of species richness for live trees at the gamma (regional) level 
    + *nStands*: number of non-zero stands (conditions) used to compute alpha-level diversity estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

<br>

## Trees Per Acre (TPA) and Basal Area
`tpa` produces tree per acre (TPA) and basal area per acre (BAA) estimates.

### **Individual-level**

To return individual tree records per year at the sub-plot level:

``` {r raw_tpa,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
trees<- tpa(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT,SUBP,STATUSCD,TREE, DIA), bySpecies = TRUE, bySizeClass =TRUE, method = "annual")
``` 

This returns a large `data.frame` with each row corresponding to an individual tree. Returns DBH and host status; see File structure for data column descriptions. This `data.frame` can then be exported using any efficient export funtion (`write.csv())`).

### **Plot-level**

To estimate TPA and Biomass of trees (>10 in. DBH) at the plot-level:

``` {r pl_tpa,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
tpaSppPlot <- tpa(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), bySpecies = TRUE,  method = "annual")
```


To calcuate by size class include argument `bySizeClass =TRUE`


### **Ecoregion-level** 

To return estimates of live trees per acre (DBH >10 in.) and basal area by Ecoregions intersecting APPA HUC10 Shell:

``` {r eco_tpa,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
tpaS <- tpa(at_FIA, polys = eco, bySpecies = TRUE, treeType = 'live', treeDomain = DIA >= 10, nCores = cores, method = 'annual')
```


### **Data definitions**

* **Trees Per Acre (TPA) and Basal Area**: Estimates of live tree abundance (TPA & BAA) per forested acre 
    + *YEAR*: FIA reporting year for current abundance estimates
    + *TPA*: estimate of trees per acre
    + *BAA*: estimate of tree basal area (sq.ft.) per acre
    + *TPA_PERC*: estimate of proportion of trees which are live, with respect to TPA
    + *BAA_PERC*: estimate of proportion of trees which are live, with respect to BAA
    + *nPlots_TREE*: number of non-zero plots used to compute tree and basal area estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Regeneration (Stems > 1 in. DBH)
`tpa` can also be used to estimate regeneration by setting the `treeDomain` argument to an appropriate size class distribution.

### **Plot-level**

To estimate regeneration per species (<5 in. DBH) at the plot-level

``` {r pl_regen,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
regenS <- tpa(at_FIA, polys=eco, byPlot = TRUE, bySpecies = TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),treeType = 'live', treeDomain = DIA < 5, nCores = cores, method = 'annual')
```

### **Ecoregion-level**

To estimate regeneration per species (<5 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_regen,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
regen <- tpa(at_FIA, polys = eco, treeType = 'live', treeDomain = DIA < 5, nCores = cores, method = 'annual')
```

<br>

### **Data Definitions**

* **Regeneration**: Estimates of live sapling (2.5 cm <= DBH < 12.7 cm) abundance (TPA & BAA) per forested acre 
    + *YEAR*: FIA reporting year for current abundance estimates
    + *TPA*: estimate of saplings per acre
    + *BAA*: estimate of sapling basal area (sq.ft.) per acre
    + *TPA_PERC*: estimate of proportion of saplings which are live, with respect to TPA
    + *BAA_PERC*: estimate of proportion of saplings which are live, with respect to BAA
    + *nPlots_TREE*: number of non-zero plots used to compute tree and basal area estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

## Mortality and Recruitment

`growMOrt` produces estimates of annual recruitment, natural mortality, and harvest rates.

Recruitment events are defined as when a live stem which is less than 10 inches DBH at time 1, grows to or beyond 10 inches DBH by time 2. This does NOT include stems which grow beyond the 10-inch diameter criteria and are then subject to mortality prior to remeasurement. Natural mortality is defined as when a live stem is subject to non-harvest mortality between successive measurement periods. Finally, harvest is defined as when a live stem is cut and removed between successive measurements.

### **Plot-level**

To estimate plot-level growth and mortality per species (trees DBH > 10 in.)

``` {r pl_gMort,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
gmSPlot <- growMort(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT),  method = "annual", treeDomain = DIA >= 10)
```

To calcuate by size class include argument `bySizeClass = TRUE`


### **Ecoregion-level**

To estimate growth and mortality per species (trees DBH > 10 in.) by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_gMort,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
gm <- growMort(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')
``` 

### **Data Definitions**

* **Mortality and Recruitment**: Estimates of annual tree (DBH >= 10 in. DBH) recruitment, mortality, and harvest rates per forested acre and relative to population totals (e.g. % mortality / year)
    + *YEAR*: FIA reporting year for growth, removal, and mortality estimates
    + *RECR_TPA*: estimate of annual recruitment (growth beyond 10 in. DBH) as trees per forested acre
    + *MORT_TPA*: estimate of annual mortality (excluding harvest and conversion) as trees per forested acre
    + *REMV_TPA*: estimate of annual removals (harvest and conversion) as trees per forested acre
    + *RECR_PERC*: estimate of annual recruitment rate, as % of individuals recruiting relative to total population
    + *MORT_PERC*: estimate of annual mortality rate, as % of individuals subject to mortality relative to total population
    + *REMV_PERC*: estimate of annual removal rate, as % of individuals subject to removal relative to total population
    + *nPlots_TREE*: number of non-zero plots used to compute total tree estimates
    + *nPlots_RECR*: number of non-zero plots used to compute recruitment estimates  
    + *nPlots_MORT*: number of non-zero plots used to compute mortality estimates    
    + *nPlots_REMV*: number of non-zero plots used to compute removal estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    

<br>

## Growth rates
`vitalRates` provides estimates of average annual DBH (inches/ yr), basal area (sq. ft./ yr), biomass (short tons/ yr), and net volume (cu. ft./yr) growth rates for individual stems, along with average annual basal area and net volume growth per acre.


### **Plot-level**

To estimate rates per species (>10 in. DBH) at the plot-level.

``` {r pl_vit,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
vrS  <- vitalRates(at_FIA, byPlot = TRUE, bySpecies = TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```


To calcuate by size class include argument `bySizeClass =TRUE`


### **Ecoregion-level**

Estimate rates per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_vit,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE }
vr  <- vitalRates(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```

### **Data Definitions**

* **Growth rates**: Estimates of (1) individual live tree diameter, basal area, net volume, and biomass growth rates, and (2) basal area, net volume, and biomass growth rates expressed on a per forested acre basis
    + *YEAR*: FIA reporting year for growth, removal, and mortality estimates
    + *DIA_GROW*: estimate of annual diameter growth (inches) rate for live trees (individual growth)
    + *BA_GROW*: estimate of annual basal area growth (sq.ft.) rate for live trees (individual growth)
    + *NETVOL_GROW*: estimate of annual net volume growth (cu.ft.) rate for live trees (individual growth)
    + *BIO_GROW*: estimate of annual biomass growth (tons) rate for live trees (individual growth)
    + *BAA_GROW*: estimate of annual basal area growth (sq.ft.) per forested acre (stand-level growth)
    + *NETVOL_GROW_AC*: estimate of annual net volume growth (cu.ft.) per forested acre (stand-level growth)
    + *BIO_GROW_AC*: estimate of annual biomass growth (tons) per forested acre (stand-level growth)
    + *nPlots_TREE*: number of non-zero plots used to compute tree estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`

<br>

## Biomass
`biomass` produces estimates of volume (cu.ft./acre), biomass (tons/acre), and carbon (tons/acre).

### **Individual-level**

To return biomass metrics per stem

```{r ind_bio,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bioS<-biomass(at_FIA, byPlot= TRUE, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT,SUBP,STATUSCD,TREE, DIA), bySpecies = TRUE, bySizeClass =TRUE, method = "annual")
```

### **Plot-level**

To estimate biomass per species (>10 in. DBH) at the plot-level

``` {r pl_bio,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bioS <- biomass(at_FIA, byPLot= TRUE, bySpecies = TRUE,grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```

### **Ecoregion-level**

To estimate biomass per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_bio,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
bio <- biomass(at_FIA, polys = eco, nCores = cores, treeDomain = DIA >= 10, method = 'annual')
```

### **Data Definitions**

* **Biomass**
  + Estimates of live tree biomass, volume, and carbon on a per forested acre basis 
    + *YEAR*: FIA reporting year for current abundance estimates
    + *NETVOL_ACRE*: estimate of live tree net volume (cu.ft.) per acre
    + *SAWVOL_ACRE*: estimate of live tree merchantable saw volume (cu.ft.) per acre
    + *BIO_AG_ACRE*: estimate of live tree aboveground biomass (tons) per acre
    + *BIO_BG_ACRE*: estimate of live tree belowground biomass (tons) per acre
    + *BIO_ACRE*: estimate of live tree total (AG + BG) biomass (tons) per acre
    + *CARB_AG_ACRE*: estimate of live tree aboveground carbon (tons) per acre
    + *CARB_BG_ACRE*: estimate of live tree belowground carbon (tons) per acre
    + *CARB_ACRE*: estimate of live tree total (AG + BG) carbon (tons) per acre
    + *nPlots_VOL*: number of non-zero plots used to compute volume, biomass, and carbon estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Snags (dead trees)
`tpa` and `biomass` can also be used to estimate snag density, volume, and biomass by setting the `treeType` argument to `'dead'`.

### **Plot-level**

To estimate snag density (trees/acre) per species (>10 in. DBH) at the plot-level

``` {r pl_snagD,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagDPlot <- tpa(at_FIA, treeType = 'dead', byPlot = TRUE,  treeDomain = DIA >= 10, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```

To estimate snag volume (cu.ft./acre) and biomass (tons/acre) per species (>10 in. DBH) at the plot-level

```{r pl_snagV,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagVPlot <- biomass(at_FIA, treeType = 'dead', byPlot = TRUE,  treeDomain = DIA >= 10, grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```

### **Ecoregion-level**

To estimate snag density (trees/acre) per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_snagD,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagD <- tpa(at_FIA, treeType = 'dead', polys = eco, treeDomain = DIA >= 10, nCores = cores, method = 'annual')
```

To estimate snag volume (cu.ft./acre) and biomass (tons/acre) per species (>10 in. DBH) by Ecoregions intersecting APPA HUC10 Shell

```{r eco_snagV,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
snagV <- biomass(at_FIA, treeType = 'dead', polys = eco, treeDomain = DIA >= 10, nCores = cores, method = 'annual')
```

### **Data Definitions**

* **Snag density and basal area**: Estimates of standing dead tree abundance (TPA & BAA) per forested acre and relative to population totals (e.g. percent snags by TPA)
    + *YEAR*: FIA reporting year for current abundance estimates
    + *TPA*: estimate of snags per acre
    + *BAA*: estimate of snags basal area (sq.ft.) per acre
    + *TPA_PERC*: estimate of proportion of snags relative to all stems (live and dead), with respect to TPA
    + *BAA_PERC*: estimate of proportion of snags relative to all stems (live and dead), with respect to BAA
    + *nPlots_TREE*: number of non-zero plots used to compute tree and basal area estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
* **Snag Volume**: Estimates of standing dead tree (DBH >= 10 in.) biomass, volume, and carbon on a per forested acre basis 
    + *YEAR*: FIA reporting year for current abundance estimates
    + *NETVOL_ACRE*: estimate of snag net volume (cu.ft.) per acre
    + *SAWVOL_ACRE*: estimate of snag merchantable saw volume (cu.ft.) per acre
    + *BIO_AG_ACRE*: estimate of snag aboveground biomass (tons) per acre
    + *BIO_BG_ACRE*: estimate of snag belowground biomass (tons) per acre
    + *BIO_ACRE*: estimate of snag total (AG + BG) biomass (tons) per acre
    + *CARB_AG_ACRE*: estimate of snag aboveground carbon (tons) per acre
    + *CARB_BG_ACRE*: estimate of snag belowground carbon (tons) per acre
    + *CARB_ACRE*: estimate of snag total (AG + BG) carbon (tons) per acre
    + *nPlots_VOL*: number of non-zero plots used to compute volume, biomass, and carbon estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
* **Stand Structure**: Estimates of forest structural stage distributions as percent forested land area in pole, mature, late, and mosaic stages
    + *YEAR*: FIA reporting year for current abundance estimates
    + *POLE_PERC*: estimate of proportion forested area in pole stage forest
    + *MATURE_PERC*: estimate of proportion forested area in mature stage forest
    + *LATE_PERC*: estimate of proportion forested area in late-seral stage forest
    + *MOSAIC_PERC*: estimate of proportion forested area in mosaic stage (unclassified) forest
    + *nPlots*: number of non-zero plots used to compute current area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Down woody debris
`dwm` produces estimates of down woody material volume (cu.ft/acre) and biomass (tons/acre). As defined by FIA, down woody material includes dead organic materials (resulting from plant mortality and leaf turnover) and fuel complexes of live shrubs and herbs. To maintain relevance for forest fuels management, we report estimates grouped by fuel lag-time classes. See Woodall et al. (2007) for definitions of fuel lag-time classes.

### **Plot-level**

To estimate down woody debris volume (cu.ft/acre) and biomass (tons/acre) at the plot-level

```{r pl_dwm,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
dwPlot <- dwm(at_FIA, byPlot = TRUE,grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```


### **Ecoregion-level**

To estimate down woody debris volume (cu.ft/acre) and biomass (tons/acre) by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_dwm, include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
dw <- dwm(at_FIA, polys = eco, tidy = FALSE, nCores = cores, method = 'annual')
``` 

### **Data Definitions**
* **Down woody material**: Estimates of down woody material volume, biomass, and carbon by fuel type on a per forested acre basis
    + *YEAR*: FIA reporting year for current abundance estimates
    + *VOL_DUFF_ACRE*: Not available for this region
    + *VOL_LITTER_ACRE*: Not available for this region
    + *VOL_1HR_ACRE*: estimate of 1 HR (small fine woody debris) volume (cu.ft.) per forested acre
    + *VOL_10HR_ACRE*: estimate of 10 HR (medium fine woody debris) volume (cu.ft.) per forested acre
    + *VOL_100HR_ACRE*: estimate of 100 HR (large fine woody debris) volume (cu.ft.) per forested acre
    + *VOL_1000HR_ACRE*: estimate of 1000 HR (coarse woody debris) volume (cu.ft.) per forested acre
    + *VOL_PILE_ACRE*: estimate of slash pile volume (cu.ft.) per forested acre
    + *VOL_ACRE*: estimate of total down woody debris (duff + litter + 1HR + 10HR + 100HR + 1000HR + pile) volume (cu.ft.) per forested acre
    + *BIO_DUFF_ACRE*: estimate of duff biomass (tons) per forested acre
    + *BIO_LITTER_ACRE*: estimate of litter biomass (tons) per forested acre
    + *BIO_1HR_ACRE*: estimate of 1 HR (small fine woody debris) biomass (tons) per forested acre
    + *BIO_10HR_ACRE*: estimate of 10 HR (medium fine woody debris) biomass (tons) per forested acre
    + *BIO_100HR_ACRE*: estimate of 100 HR (large fine woody debris) biomass (tons) per forested acre
    + *BIO_1000HR_ACRE*: estimate of 1000 HR (coarse woody debris) biomass (tons) per forested acre
    + *BIO_PILE_ACRE*: estimate of slash pile biomass (tons) per forested acre
    + *BIO_ACRE*: estimate of total down woody debris (duff + litter + 1HR + 10HR + 100HR + 1000HR + pile) biomass (tons) per forested acre
    + *CARB_DUFF_ACRE*: estimate of duff carbon (tons) per forested acre
    + *CARB_LITTER_ACRE*: estimate of litter carbon (tons) per forested acre
    + *CARB_1HR_ACRE*: estimate of 1 HR (small fine woody debris) carbon (tons) per forested acre
    + *CARB_10HR_ACRE*: estimate of 10 HR (medium fine woody debris) carbon (tons) per forested acre
    + *CARB_100HR_ACRE*: estimate of 100 HR (large fine woody debris) carbon (tons) per forested acre
    + *CARB_1000HR_ACRE*: estimate of 1000 HR (coarse woody debris) carbon (tons) per forested acre
    + *CARB_PILE_ACRE*: estimate of slash pile carbon (tons) per forested acre
    + *CARB_ACRE*: estimate of total down woody debris (duff + litter + 1HR + 10HR + 100HR + 1000HR + pile) carbon (tons) per forested acre
    + *nPlots*: number of non-zero plots used to compute down woody material abundance and area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
<br>

## Invasive plant abundance
`invasive` produces estimates of the areal coverage (%) of invasive species and frequency of plots invasives species were detected in.

Estimates of areal coverage of individual invasive species should NOT be summed to produce estimates of areal coverage by ALL invasive species, as areal coverage by species is not mutually exclusive (multiple species my occur in the same area). Current FIA data collection protocols do not allow for the unbiased estimation of areal coverage by all invasive species.

### **Plot-level**

To estimate invasive plant cover per species at the plot-level

```{r pl_inv,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
invPlot <- invasive(at_FIA, byPlot = TRUE,grpBy = c(STATECD, ECOSUBCD, OWNGRPCD, QA_STATUS,PLOT), nCores = cores, method = 'annual')
```


### **Ecoregion-level**
While the invasive data are fairly sparse (as of 2020) in the FIADB, you can estimate invasive plant cover per species by Ecoregions intersecting APPA HUC10 Shell

``` {r eco_inv,include = TRUE, echo=TRUE, warning=FALSE, comment= NA, eval= FALSE}
inv <- invasive(at_FIA, polys = eco, nCores = cores, method = 'annual')
```
### **Data Definitions**

* **Invasive Species Cover**: Estimates of areal coverage by invasive plant species relative to forested land area
    + *YEAR*: FIA reporting year for current abundance estimates
    + *SYMBOL*: unique species ID from NRCS Plant Reference Guide
    + *SCIENTIFIC_NAME*: latin name of the species
    + *COMMON_NAME*: commmon name of the species
    + *COVER_PCT*: estimate of percent areal coverage with respect to total forested land area
    + *nPlots_INV*: number of non-zero plots used to compute species coverage estimates
    + *nPlots_AREA*: number of non-zero plots used to compute land area estimates
    + columns ending in *SE*: estimates of sampling error (%) of the respective variable. All sampling error estimates are computed with 68% confidence
    + other columns from `at_ecoSub.shp`
    
# References

Bechtold, W.A.; Patterson, P.L., eds. 2005. The Enhanced Forest Inventory and Analysis Program - National Sampling Design and Estimation Procedures. Gen. Tech. Rep. SRS - 80. Asheville, NC: U.S. Department of Agriculture, Forest Service, Southern Research Station. 85 p. https://www.srs.fs.usda.gov/pubs/gtr/gtr_srs080/gtr_srs080.pdf

Dieffenbach, F. (2018). Appalachian National Scenic Trail forest health monitoring protocol, Natural Resource Report NPS/NETN/NRR—2018/1804. National Park Service, Fort Collins, Colorado. https://irma.nps.gov/DataStore/DownloadFile/610353

Woodall, C.; Monleon, V.J., eds. 2007. Sampling Protocol, Estimation, and Analysis Procedures for the Down Woody Materials Indicator of the FIA Program. Gen. Tech. Rep. NRS - 22. ewtown Square, PA: U.S. Department of Agriculture, Forest Service, Northern Research Station. https://www.nrs.fs.fed.us/pubs/gtr/gtr_nrs22.pdf

FIA Database User Guide: https://www.fia.fs.fed.us/library/database-documentation/

# Document versioning 

This document was created using R version `r getRversion()` on `r Sys.Date()` by Aaron Weed.
